Biogeography of crop progenitors and wild plant resources over the Pleistocene/Holocene boundary in West Asia, 14.7--8.3 ka

### 2025-01-05 analysis/paper.qmd
Add discussion of comparison with arch data
Removing LGM and HS1 as we don't have sites from them

---
title: Biogeography of crop progenitors and wild plant resources in the Late Pleistocene and Early Holocene of West Asia, 21–8.3 ka
date: today
author: 
- name: Joe Roe
  id: jr
  orcid: 0000-0002-1011-1244
  email: joeroe@hey.com
  affiliations:
    - name: University of Bern
      country: Switzerland
    - name: University of Copenhagen
      country: Denmark
  corresponding: true
- name: Amaia Arranz-Otaegui
  id: aao
  orcid: 0000-0002-5091-6426
  affiliation: 
    - name: University of the Basque Country
      country: Spain
abstract: |
  This paper presents the first comprehensive reconstructions of the 
  palaeodistributions of 56 plant species found regularly in association with
  early agricultural archaeological sites in West Asia, including the 
  progenitors of the first crops. We used machine learning to train an 
  ecological niche model of each species based on its present-day distribution 
  in relation to climate and environmental variables. Predictions of the 
  potential ranges of these species at key stages of the Pleistocene–Holocene 
  transition could then be derived from these models using hindcast data from 
  palaeoclimate simulations.
  Species ranges are predicted to have been on average [X%] [larger|smaller] in 
  the Early Holocene compared to present conditions, but did [not] change as
  drastically between the Last Glacial Maximum and the Early Holocene. 
  However, the modelled ranges do not predict the observed occurrence of these 
  species on archaeological sites with any degree of accuracy. The regional 
  ubiquity of species in the archaeological record is [not] correlated with the 
  predicted size of its range and the diversity of archaeobotanical assemblages is [not] 
  correlated with the predicted diversity of its environs. This indicates that 
  trends in taxonomic composition of the archaeobotanical record is [not] likely 
  to have been influenced by environmental change and species turnover, in
  addition to human economic choices.
format: 
  elsevier-pdf:
    keep-tex: true
    # cite-method: citeproc
    journal:
      name: Open Quaternary
      formatting: review
      cite-style: authoryear
bibliography: references.bib
execute: 
  echo: false
---

* [x] Finish modelling
- [ ] First draft:
  - [ ] Introduction
  - [ ] Background
    - [ ] Biogeography
    - [x] ENM
  - [x] Methods & Materials
  - [ ] Results
    - [ ] Model assessment
  - [ ] Discussion
    - [ ] General trends
    - [ ] Case studies
  - [ ] Conclusion
- [ ] Figures
  - [ ] tbl-occ-count (needs pdf cleanup)
  - [ ] tbl-predictors
  - [ ] tbl-climate-periods
- [ ] Clean up package code
- [ ] Appendix with all hindcast predictions
- [ ] References & copyediting
- [ ] Final proofread

@@

# TODO: drop LGM and HS1 since there are no sites 
# (can then maybe increase prediction resolution?)
climate_periods <- tibble(
  period = c("Last Glacial Maximum", "Heinrich Stadial 1", "Bølling-Allerød",
             "Younger Dryas", "Early Holocene", "Current"),
  code = c("lgm", "hs1", "ba", "yds", "eh", "cur"),
  label = fct_inorder(c("LGM (21 ka)", "HS1 (17.0-14.7 ka)", "BA (14.7-12.9 ka)", 
            "YDS (12.9-11.7 ka)", "EH (11.7-8.3 ka)", "current")),
  start_bp = c(21000, 17000, 14700, 12900, 11700, NA),
  end_bp = c(17000, 14700, 12900, 11700, 8300, NA)
)
```

@@

* Has always been important in study of agricultural origins
  * Historically: Vavilov, Pumpelly & Childe
  * Genetic studies tell us origin points, but not ranges
* Important to e.g.
  * Distinguish environmental from potentially anthropogenic change [@MartinEtAl2017; @MartinEtAl2025]
  * Reconstruct sequences of domestication [@YeomansEtAl2017]
* Epipal./Neo. plant-based economies were diverse
  * More than the "founder crops"; 
  * More than food
  * (In archaebot., not all intentionally collected)
  * Regionally and temporally diverse
  * ...so we model lots of species!
* Regional ecological reconstructions generally rely on the 'expert interpolation' (or what do they call it with isoscapes?) method
  * See CSEAS (AEA-prep) presentation
  * Figure: comparisons

@@

ne_land <- read_sf(raw_data("ne"), "ne_10m_land") |> 
  st_geometry() |>
  st_crop(buffer_bbox(w_asia, 10)) |>
  st_difference(ne_ocean) |>
  st_difference(ne_lakes) |>
  st_union()

@@

```{r fig-region}
#| fig-cap: Study region (red) and archaeological sites used to generate modelled flora. Solid circles indicate Neolithic sites.
# TODO: better pallette
ggplot() + 
  annotation_spatial(ne_land, fill = "#ffffff") +
  annotation_spatial(ne_rivers, aes(linewidth = strokeweig), colour = "blue") +
  annotation_spatial(ne_lakes, colour = "blue", fill = "lightblue") +
  layer_spatial(w_asia, colour = "red", fill = NA) +
  layer_spatial(archaeo_sites, shape = 1) +
  layer_spatial(filter(archaeo_sites, period %in% neolithic), shape = 16) +
  scale_linewidth_identity() +
  coord_sf(crs = w_asia_albers)
```

@@

```{r filter-flora}
flora <- filter(flora, n_occ > min_n_occ)
# Why are there 50k occurrences of Avena sterilis??
flora <- filter(flora, taxon != "Avena sterilis")
```

```{r fetch-occ}
# TODO: further tidying. Ideas:
# - https://data-blog.gbif.org/post/gbif-filtering-guide/
# - https://cran.r-project.org/web/packages/CoordinateCleaner/vignettes/Cleaning_GBIF_data_with_CoordinateCleaner.html
# - https://www.r-bloggers.com/2021/03/downloading-and-cleaning-gbif-data-with-r/
if (!file_exists(derived_data("flora_occ.tsv"))) {
  message("Reacquiring GBIF occurrences...")
  flora <- mutate(
    flora,
    occ = map(
      taxon, gbif_data, 
      hasCoordinate = TRUE, hasGeospatialIssue = FALSE, geometry = w_asia, 
      limit = 100000, .progress = TRUE
    ),
    occ = map(occ, gbif_drop_imprecise_coords, threshold = 5000),
    occ = map(occ, gbif_drop_fossils),
    occ = map(occ, gbif_drop_duplicate_coords),
    occ = map(occ, gbif_standardise_names),
    occ = map(occ, \(x) {
      select(x, gbif_key = key, scientific_name, genus, species, longitude, 
             latitude)
    })
  )
  write_tsv(unnest(flora, cols = c(occ)), derived_data("flora_occ.tsv"))
}
flora <- read_tsv(
  derived_data("flora_occ.tsv"),
  col_types = cols(
    taxon = col_character(),
    taxon_detail = col_character(),
    n_present = col_double(),
    p_present = col_double(),
    n_occ = col_double(),
    gbif_key = col_double(),
    scientific_name = col_character(),
    genus = col_character(),
    species = col_character(),
    longitude = col_double(),
    latitude = col_double()
  )
) |>
  nest(occ = gbif_key:latitude)
```

@@

```{r tbl-climate-periods}
#| fig-cap: Palaeoclimatic periods used for hindcasting, after @BrownEtAl2018
# TODO: add refs (as footnotes?)
climate_periods |>
  mutate(
    code = toupper(code),
    age_start = start_bp / 1000,
    age_end = end_bp / 1000
  ) |>
  select(period, code, age_start, age_end) |>
  gt() |>
  cols_label(period = "Period", age_start = "Age") |>
  cols_merge(
    c(period, code),
    pattern = "{1} ({2})"
  ) |>
  cols_merge_range(age_start, age_end) |>
  cols_units(age_start = "ka") |>
  sub_missing()
```

@@

```{r fit-niche-models}
# Fit best model to training data
if (!file_exists(derived_data("flora_niche_fit.Rdata"))) {
  flora <- mutate(
    flora,
    niche = map(occ_split, \(x) last_fit(niche_workflow, x), .progress = TRUE)
  )
  saveRDS(flora, derived_data("flora_niche_fit.Rdata"))
}
flora <- readRDS(derived_data("flora_niche_fit.Rdata"))
```

@@

```{r predict}
# NOTE: very high memory usage
if (!file_exists(derived_data("flora_niche_predictions.Rdata"))) {
  flora <- flora |>
    mutate(
      prediction = map(niche, \(mod, obj) predict(obj, extract_fit_engine(mod)),
                       obj = environment, .progress = TRUE)
    )
  saveRDS(flora, derived_data("flora_niche_predictions.Rdata"))
}
flora <- readRDS(derived_data("flora_niche_predictions.Rdata"))
```

@@

```{r data-model-assessment}
flora <- mutate(
  flora,
  niche_test_metrics = map(niche, collect_metrics)
)

@@

```{r fig-model-accuracy-vs-n_occ}
#| fig-cap: Number of taxon occurrences and model accuracy on test dataset
flora |>
  unnest(niche_test_metrics) |>
  filter(.metric == "accuracy") |>
  mutate(label = if_else(.estimate < 0.80, taxon, NA_character_)) |>
  ggplot(aes(n_occ, .estimate)) + 
  geom_smooth(method = "glm", formula = y ~ log(x)) +
  geom_point() + 
  geom_text_repel(aes(label = label), fontface = "italic") +
  scale_x_log10(name = "Occurrences") +
  scale_y_continuous(name = "Model accuracy", labels = label_percent()) +
  theme_minimal_grid() 
```

We assessed the predictive performance of the fitted niche models in the contemporary environment based on the reserved test partition.
Model accuracy (i.e. proportion of correctly classified test samples) ranged between `r percent(filter(niche_test_metrics, .metric == "accuracy")$min)` and `r percent(filter(niche_test_metrics, .metric == "accuracy")$max)`, with an average of `r percent(filter(niche_test_metrics, .metric == "accuracy")$mean)`.
The area under the models' receiver operating characteristic curves (ROC-AUC) was in the range of `r round(filter(niche_test_metrics, .metric == "roc_auc")$mean, 3)`±`r round(filter(niche_test_metrics, .metric == "accuracy")$sd, 3)`
Model accuracy is loosely correlated with the number of occurrences available for training (@fig-model-accuracy-vs-n_occ), though the worst-performing models also included some species with relatively large sample sizes: <!-- TODO e.g. -->.
Test metrics and ROC curves for the individual models are included in the appendix.

@@

* General trends:
  * Quantified sensitivity of plant ranges to climate change
  * Crop progenitors saw range contractions just before the onset of agriculture? (Moreso than other wild resources??)
  * How could are reconstructed ranges at predicting archaeological assemblages? (+Implications)
* Interesting individual case studies:
  * Wheat progenitors
  * ???

@@

>>>>

---
title: Biogeography of crop progenitors and wild plant resources over in terminal Pleistocene and Early Holocene boundary West Asia, 14.7–8.3 ka
date: today
author: 
- name: Joe Roe
  id: jr
  orcid: 0000-0002-1011-1244
  email: joeroe@hey.com
  affiliations:
    - name: University of Bern
      country: Switzerland
    - name: University of Copenhagen
      country: Denmark
  corresponding: true
- name: Amaia Arranz-Otaegui
  id: aao
  orcid: 0000-0002-5091-6426
  affiliation: 
    - name: University of the Basque Country
      country: Spain
abstract: |
  This paper presents the first comprehensive reconstructions of the 
  palaeodistributions of 56 plant species found regularly in association with
  early agricultural archaeological sites in West Asia, including the 
  progenitors of the first crops. We used machine learning to train an 
  ecological niche model of each species based on its present-day distribution 
  in relation to climate and environmental variables. Predictions of the 
  potential ranges of these species at key stages of the Pleistocene–Holocene 
  transition could then be derived from these models using hindcast data from 
  palaeoclimate simulations.
  Species ranges are predicted to have been on average [X%] [larger|smaller] in 
  the Early Holocene compared to present conditions, but did [not] change as
  drastically between the Last Glacial Maximum and the Early Holocene. 
  However, the modelled ranges do not predict the observed occurrence of these 
  species on archaeological sites with any degree of accuracy. The regional 
  ubiquity of species in the archaeological record is [not] correlated with the 
  predicted size of its range and the diversity of archaeobotanical assemblages is [not] 
  correlated with the predicted diversity of its environs. This indicates that 
  trends in taxonomic composition of the archaeobotanical record is [not] likely 
  to have been influenced by environmental change and species turnover, in
  addition to human economic choices.
format: 
  elsevier-pdf:
    keep-tex: true
    # cite-method: citeproc
    journal:
      name: Open Quaternary
      formatting: review
      cite-style: authoryear
bibliography: references.bib
execute: 
  echo: false
---

* [x] Finish modelling
- [ ] First draft:
  - [ ] Introduction
  - [ ] Background
    - [ ] Biogeography
    - [x] ENM
  - [x] Methods & Materials
  - [x] Results
  - [ ] Discussion
    - [ ] General trends
    - [ ] Case studies
    - [x] Comparison with archaeobot
  - [ ] Conclusion
- [ ] Figures
  - [x] fig-region
  - [ ] tbl-occ-count (needs pdf cleanup)
  - [ ] tbl-predictors
- [ ] Outstanding TODOs
- [ ] Appendix with all hindcast predictions
- [ ] Copyedit and finalise references
- [ ] Clean up package code
- [ ] Final proofread

@@

climate_periods <- tibble(
  period = c("Bølling-Allerød", "Younger Dryas", "Early Holocene", "Current"),
  code = c("ba", "yds", "eh", "cur"),
  label = fct_inorder(c("BA (14.7-12.9 ka)", "YDS (12.9-11.7 ka)", 
                      "EH (11.7-8.3 ka)", "current")),
  start_bp = c(14700, 12900, 11700, NA),
  end_bp = c(12900, 11700, 8300, NA)
)
```

@@

* Has always been important in study of agricultural origins
  * Historically: Vavilov, Pumpelly & Childe
  * Genetic studies tell us origin points, but not ranges
* Important to e.g.
  * Distinguish environmental from potentially anthropogenic change [@MartinEtAl2017; @MartinEtAl2025]
  * Reconstruct sequences of domestication [@YeomansEtAl2017]
* Epipal./Neo. plant-based economies were diverse
  * More than the "founder crops"; 
  * More than food
  * (In archaebot., not all intentionally collected)
  * Geographically and temporally diverse
  * ...so we model lots of species!
* Regional ecological reconstructions generally rely on the 'expert interpolation' (or what do they call it with isoscapes?) method
  * See CSEAS (AEA-prep) presentation
  * Figure: comparisons

@@

ne_land <- read_sf(raw_data("ne"), "ne_10m_land") |> 
  st_geometry() |>
  st_crop(buffer_bbox(w_asia, 10)) |>
  st_difference(ne_ocean) |>
  # st_difference(ne_lakes) |>
  st_union()

@@

```{r fig-region}
#| fig-cap: Study region (red) and archaeological sites used to generate modelled flora. Solid circles indicate Neolithic sites.
ggplot() + 
  annotation_spatial(ne_land, fill = "#ffffff") +
  annotation_spatial(ne_rivers, aes(linewidth = strokeweig), colour = "lightblue") +
  layer_spatial(w_asia, colour = "red", fill = NA) +
  layer_spatial(archaeo_sites, shape = 1) +
  layer_spatial(filter(archaeo_sites, period %in% neolithic), shape = 16) +
  scale_linewidth_identity() +
  coord_sf(crs = w_asia_albers)
```

@@

```{r data-filter-flora}
flora <- filter(flora, n_occ > min_n_occ)
# Why are there 50k occurrences of Avena sterilis??
flora <- filter(flora, taxon != "Avena sterilis")
```

```{r data-occ}
# TODO: further tidying. Ideas:
# - https://data-blog.gbif.org/post/gbif-filtering-guide/
# - https://cran.r-project.org/web/packages/CoordinateCleaner/vignettes/Cleaning_GBIF_data_with_CoordinateCleaner.html
# - https://www.r-bloggers.com/2021/03/downloading-and-cleaning-gbif-data-with-r/
if (!file_exists(derived_data("flora_occ.tsv"))) {
  message("Reacquiring GBIF occurrences...")
  flora <- mutate(
    flora,
    occ = map(
      taxon, gbif_data, 
      hasCoordinate = TRUE, hasGeospatialIssue = FALSE, geometry = w_asia, 
      limit = 100000, .progress = TRUE
    ),
    occ = map(occ, gbif_drop_imprecise_coords, threshold = 5000),
    occ = map(occ, gbif_drop_fossils),
    occ = map(occ, gbif_drop_duplicate_coords),
    occ = map(occ, gbif_standardise_names),
    occ = map(occ, \(x) {
      select(x, gbif_key = key, scientific_name, genus, species, longitude, 
             latitude)
    })
  )
  write_tsv(unnest(flora, cols = c(occ)), derived_data("flora_occ.tsv"))
}
flora <- read_tsv(
  derived_data("flora_occ.tsv"),
  col_types = cols(
    taxon = col_character(),
    taxon_detail = col_character(),
    n_present = col_double(),
    p_present = col_double(),
    n_occ = col_double(),
    gbif_key = col_double(),
    scientific_name = col_character(),
    genus = col_character(),
    species = col_character(),
    longitude = col_double(),
    latitude = col_double()
  )
) |>
  nest(occ = gbif_key:latitude)
```

@@

```{r tbl-climate-periods}
#| fig-cap: Palaeoclimatic periods used for hindcasting, after @BrownEtAl2018
# TODO: integrate into text
climate_periods |>
  mutate(
    code = toupper(code),
    age_start = start_bp / 1000,
    age_end = end_bp / 1000
  ) |>
  select(period, code, age_start, age_end) |>
  gt() |>
  cols_label(period = "Period", age_start = "Age") |>
  cols_merge(
    c(period, code),
    pattern = "{1} ({2})"
  ) |>
  cols_merge_range(age_start, age_end) |>
  cols_units(age_start = "ka") |>
  sub_missing()
```

@@

```{r niche-fit}
# Fit best model to training data
if (!file_exists(derived_data("flora_niche_fit.Rdata"))) {
  flora <- mutate(
    flora,
    niche = map(occ_split, \(x) last_fit(niche_workflow, x), .progress = TRUE)
  )
  saveRDS(flora, derived_data("flora_niche_fit.Rdata"))
}
flora <- readRDS(derived_data("flora_niche_fit.Rdata"))
```

@@

```{r niche-predict}
# NOTE: high memory usage
if (!file_exists(derived_data("flora_niche_predictions.Rdata"))) {
  flora <- flora |>
    mutate(
      prediction = map(niche, \(mod, obj) predict(obj, extract_fit_engine(mod)),
                       obj = environment, .progress = TRUE)
    )
  saveRDS(flora, derived_data("flora_niche_predictions.Rdata"))
}
flora <- readRDS(derived_data("flora_niche_predictions.Rdata"))
```

@@

```{r data-model-assessment}
flora <- mutate(
  flora,
  pred = map(niche, collect_predictions),
  niche_test_metrics = map(niche, collect_metrics),
  sensitivity = map(pred, sens, truth = present, estimate = .pred_class,
                    event_level = "second"), # reference present=TRUE
  niche_test_metrics = map2(niche_test_metrics, sensitivity, bind_rows)
) |>
  select(-sensitivity)

@@

```{r fig-model-sensitivity-vs-n_occ}
#| fig-cap: Number of taxon occurrences and model sensitivity on test dataset
flora |>
  unnest(niche_test_metrics) |>
  filter(.metric == "sens") |>
  mutate(label = if_else(.estimate < 0.85, taxon, NA_character_)) |>
  ggplot(aes(n_occ, .estimate)) + 
  geom_smooth(method = "glm", formula = y ~ log(x)) +
  geom_point() + 
  geom_text_repel(aes(label = label), fontface = "italic") +
  scale_x_log10(name = "Occurrences") +
  scale_y_continuous(name = "Sensitivity", labels = label_percent()) +
  theme_minimal_grid() 
```

We assessed the predictive performance of the fitted niche models in the contemporary environment based on the reserved test partition.
Model accuracy (proportion of correctly classified presence and background samples) ranged between `r percent(filter(niche_test_metrics, .metric == "accuracy")$min)` and `r percent(filter(niche_test_metrics, .metric == "accuracy")$max)`, with an average of `r percent(filter(niche_test_metrics, .metric == "accuracy")$mean)`.
Sensitivity (proportion of correctly classified presence samples) ranged between `r percent(filter(niche_test_metrics, .metric == "sens")$min)` and `r percent(filter(niche_test_metrics, .metric == "sens")$max)`, with an average of `r percent(filter(niche_test_metrics, .metric == "sens")$mean)`.
The area under the models' receiver operating characteristic curves (ROC-AUC) was in the range of `r round(filter(niche_test_metrics, .metric == "roc_auc")$mean, 3)`±`r round(filter(niche_test_metrics, .metric == "accuracy")$sd, 3)`
Model sensitivity is loosely correlated with the number of occurrences available for training (@fig-model-sensitivity-vs-n_occ), with the worst-performing models all having less than 300 recorded occurrences: `r knitr::combine_words(paste0("*", filter(unnest(flora, niche_test_metrics), .metric == "sens" & .estimate < 0.85)$taxon, "*"))`.
Test metrics and ROC curves for the individual models are included in the appendix.

@@

* General trends:
  * Quantified sensitivity of plant ranges to climate change
  * Crop progenitors saw range contractions just before the onset of agriculture? (Moreso than other wild resources??)
* Interesting individual case studies:
  * Wheat progenitors
  * ???

## Comparison with archaeobotanical data

The failure of our hindcast models to predict the occurrence of species in archaeobotanical assemblages has several possible explanations.
Since they do accurately predict the test dataset <!-- TODO: and past negatives? -->, a likely culprit is overfitting of the models to the present environment.
This implies that the modelled palaeodistributions should be seen as conservative estimates or a minimal range.
Another obvious flaw in our methodology is that the time slices used for palaeoclimatic reconstruction are very broad—each covering around two millennia—and therefore potentially unrepresentative of the environment around sites at the specific time at which they were occupied.
The variable quality of the archaeological test dataset, especially in terms of chronology, is also a plausible factor.

At the same time, we cannot rule out more substantive reasons for the discrepency between predicted and observed archaeological occurrences.
The niches of the modelled species could have changed since the Early Holocene, which would not be captured in a model trained purely on modern specimens.
Human economic choices—mobility, foraging strategies, cultivation, etc.—could also produce archaeobotanical assemblages whose composition depart significantly from that of the surrounding local flora.
Further refinement of the methodology for hindcast palaeoecological niche models, for example using more finely resolved palaeoclimate sequences [e.g. @KargerEtAl2023], hyperparameter tuning to avoid overfitting, and improved archaeological datasets, would help disentangle these potential explanations.

<!-- TODO: do model predictions match more aggregated stats e.g. biodiversity? -->

@@

* We present the first comprehensive reconstructions of the palaeodistributions of 56 plant species found regularly in association with early agricultural archaeological sites in West Asia
  * R
* Modelling at scale using random forest, modern occurrences, and hindcasting represents a significant advance in pENM methodology
  * Relies on recent open ecological and climatic datasets
  * ...open archaeological datasets still lacking!
* First (? - check Yaworsky) attempt to verify hindcast models with archaeo. compositional data
  * Results not to promising, but this doesn't mean the models are useless!
  * Discrepencies suggests several areas for further research and methodological development

### 2025-01-05 analysis/paper.tex
Add discussion of comparison with arch data
Removing LGM and HS1 as we don't have sites from them

\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Biogeography of crop progenitors and wild plant resources in Late Pleistocene--Early Holocene West Asia, 21--8.3 ka},
  pdfauthor={Joe Roe; Amaia Arranz-Otaegui},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

@@

\begin{frontmatter}
\title{Biogeography of crop progenitors and wild plant resources in Late
Pleistocene--Early Holocene West Asia, 21--8.3 ka}
\author[1,2]{Joe Roe%
\corref{cor1}%
}
 \ead{joeroe@hey.com} 
\author[3]{Amaia Arranz-Otaegui%
%
}

@@

\begin{abstract}
This paper presents the first comprehensive reconstructions of the
palaeodistributions of 56 plant species found regularly in association
with early agricultural archaeological sites in West Asia, including the
progenitors of the first crops. We used machine learning to train an
ecological niche model of each species based on its present-day
distribution in relation to climate and environmental variables.
Predictions of the potential ranges of these species at key stages of
the Pleistocene--Holocene transition could then be derived from these
models using hindcast data from palaeoclimate simulations. Species
ranges are predicted to have been on average {[}X\%{]}
{[}larger\textbar smaller{]} in the Early Holocene compared to present
conditions, indicating {[}\ldots{]}. The modelled ranges predict the
observed occurrence of these species on archaeological sites with
{[}low\textbar medium\textbar high{]} accuracy. The regional ubiquity of
species in the archaeological record is {[}not{]} correlated with the
predicted size of its range and the diversity of archaeobotanical
assemblages is {[}not{]} correlated with the predicted diversity of its
environs. This indicates that trends in taxonomic composition of the
archaeobotanical record is {[}not{]} likely to have been influenced by
environmental change and species turnover, in addition to human economic
choices.
\end{abstract}

@@

    \begin{itemize}
    \tightlist
    \item[$\square$]
      Model assessment
    \item[$\square$]
      Hindcasting
    \end{itemize}
  \item[$\square$]
    Discussion

@@

\caption{\label{fig-region}Study region}

@@

\fontsize{12.0pt}{14.4pt}\selectfont
\begin{tabular*}{\linewidth}{@{\extracolsep{\fill}}lr}
\toprule
Period & Age, ka \\ 
\midrule\addlinespace[2.5pt]
Last Glacial Maximum (LGM) & 21.0–17.0 \\ 
Heinrich Stadial 1 (HS1) & 17.0–14.7 \\ 
Bølling-Allerød (BA) & 14.7–12.9 \\ 
Younger Dryas (YDS) & 12.9–11.7 \\ 
Early Holocene (EH) & 11.7–8.3 \\ 
Current (CUR) & — \\ 
\bottomrule
\end{tabular*}

@@

For hindcasting, we used reconstructed bioclimatic data for 5 key
periods (Table~\ref{tbl-climate-periods}) generated from downscaled
paleoclimate simulations from the HadCM3 general circulation model
\citep{BrownEtAl2018}. Terrain and soil predictors were held constant,
since reconstructions of these variables in the past are not available
at sufficient scale. It is not likely that either macroscale topography
or soil characteristics have altered significantly over the period of
time considered here, so we assume that this does not degrade model
performance, and may in fact benefit it by providing `anchoring'
predictors that are independent of climate change.

@@

The output of the model is probabilistic. However, this should not be
understood as an actual probability of occurrence \citep{CITE}, but more
akin to an estimate of habitat suitability. To simplify interpretation,
we can convert these predictions into binary presence/absence maps, a
process called ``thresholding''. We select the threshold value for each
model individually, using MaxSSS \citep[as recommended
by][]{LiuEtAl2013}. This also makes it possible to analyses the
predictions together as an assemblage.

@@

@@

\pandocbounded{\includegraphics[keepaspectratio]{paper_files/figure-pdf/fig-model-accuracy-vs-n_occ-1.pdf}}

@@

\caption{\label{fig-model-accuracy-vs-n_occ}Number of taxon occurrences
and model accuracy on test dataset}

@@

\begin{itemize}
\tightlist
\item
  Modelled ecological niches on current data
\end{itemize}

\subsection{Hindcasting}\label{hindcasting}

Sensitivity to climate fluctuations?

\begin{itemize}
\tightlist
\item
  Comparison to archaeological occurrences
\end{itemize}

@@

  \begin{itemize}
  \tightlist
  \item
    Quantified sensitivity of plant ranges to climate change
  \item
    Crop progenitors saw range contractions just before the onset of
    agriculture? (Moreso than other wild resources??)
  \item
    How could are reconstructed ranges at predicting archaeological
    assemblages? (+Implications)
  \end{itemize}
\item
  Interesting individual case studies:

@@

@@

\renewcommand\refname{References}
  \bibliography{references.bib}

>>>>

\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Biogeography of crop progenitors and wild plant resources over the Pleistocene/Holocene boundary in West Asia, 14.7--8.3 ka},
  pdfauthor={Joe Roe; Amaia Arranz-Otaegui},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

@@

\begin{frontmatter}
\title{Biogeography of crop progenitors and wild plant resources over
the Pleistocene/Holocene boundary in West Asia, 14.7--8.3 ka}
\author[1,2]{Joe Roe%
\corref{cor1}%
}
 \ead{joeroe@hey.com} 
\author[3]{Amaia Arranz-Otaegui%
%
}

@@

\begin{abstract}
This paper presents the first comprehensive reconstructions of the
palaeodistributions of 56 plant species found regularly in association
with early agricultural archaeological sites in West Asia, including the
progenitors of the first crops. We used machine learning to train an
ecological niche model of each species based on its present-day
distribution in relation to climate and environmental variables.
Predictions of the potential ranges of these species at key stages of
the Pleistocene--Holocene transition could then be derived from these
models using hindcast data from palaeoclimate simulations. Species
ranges are predicted to have been on average {[}X\%{]}
{[}larger\textbar smaller{]} in the Early Holocene compared to present
conditions, but did {[}not{]} change as drastically between the Last
Glacial Maximum and the Early Holocene. However, the modelled ranges do
not predict the observed occurrence of these species on archaeological
sites with any degree of accuracy. The regional ubiquity of species in
the archaeological record is {[}not{]} correlated with the predicted
size of its range and the diversity of archaeobotanical assemblages is
{[}not{]} correlated with the predicted diversity of its environs. This
indicates that trends in taxonomic composition of the archaeobotanical
record is {[}not{]} likely to have been influenced by environmental
change and species turnover, in addition to human economic choices.
\end{abstract}

@@

    \begin{itemize}
    \tightlist
    \item[$\square$]
      Model assessment
    \end{itemize}
  \item[$\square$]
    Discussion

@@

\caption{\label{fig-region}Study region (red) and archaeological sites
used to generate modelled flora. Solid circles indicate Neolithic
sites.}

@@

\fontsize{12.0pt}{14.4pt}\selectfont
\begin{tabular*}{\linewidth}{@{\extracolsep{\fill}}lr}
\toprule
Period & Age, ka \\ 
\midrule\addlinespace[2.5pt]
Bølling-Allerød (BA) & 14.7–12.9 \\ 
Younger Dryas (YDS) & 12.9–11.7 \\ 
Early Holocene (EH) & 11.7–8.3 \\ 
Current (CUR) & — \\ 
\bottomrule
\end{tabular*}

@@

For hindcasting, we used reconstructed bioclimatic data for 3 key
periods (Table~\ref{tbl-climate-periods}) generated from downscaled
paleoclimate simulations from the HadCM3 general circulation model
\citep{BrownEtAl2018}. Terrain and soil predictors were held constant,
since reconstructions of these variables in the past are not available
at sufficient scale. It is not likely that either macroscale topography
or soil characteristics have altered significantly over the period of
time considered here, so we assume that this does not degrade model
performance, and may in fact benefit it by providing `anchoring'
predictors that are independent of climate change.

@@

@@

We trained Random Forest models for 56 taxa using contemporary
occurrence data from GBIF, a random sample of background points, and the
predictor variables described in \textbf{?@tbl-predictors}. Substituting
the ``current'' climate predictors for those derived from palaeoclimatic
simulations \citep{BrownEtAl2018}, we could then generate hindcast
predictions for reconstructed past environments in 4 key climate periods
-- a total of 224 modelled palaeodistributions. Predicted distributions
for individual taxa are presented in the appendix and accompanying
material.

@@

\pandocbounded{\includegraphics[keepaspectratio]{paper_files/figure-pdf/fig-model-sensitivity-vs-n_occ-1.pdf}}

@@

\caption{\label{fig-model-sensitivity-vs-n_occ}Number of taxon
occurrences and model sensitivity on test dataset}

@@

We assessed the predictive performance of the fitted niche models in the
contemporary environment based on the reserved test partition. Model
accuracy (proportion of correctly classified presence and background
samples) ranged between 72\% and 98\%, with an average of 91\%.
Sensitivity (proportion of correctly classified presence samples) ranged
between 80\% and 100\%, with an average of 93\%. The area under the
models' receiver operating characteristic curves (ROC-AUC) was in the
range of 0.973±0.058 Model sensitivity is loosely correlated with the
number of occurrences available for training
(Figure~\ref{fig-model-sensitivity-vs-n_occ}), with the worst-performing
models all having less than 300 recorded occurrences: \emph{Buglossoides
tenuiflora}, \emph{Cephalaria syriaca}, \emph{Gypsophila pilosa},
\emph{Lathyrus sativus}, \emph{Triticum turgidum dicoccum}. Test metrics
and ROC curves for the individual models are included in the appendix.

The ability of the hindcast models to predict the occurrence of specific
species at archaeological sites is significantly worse, with only 12\%
of presences in archaeobotanical assemblages successfully predicted.

@@

  \begin{itemize}
  \tightlist
  \item
    Quantified sensitivity of plant ranges to climate change
  \item
    Crop progenitors saw range contractions just before the onset of
    agriculture? (Moreso than other wild resources??)
  \end{itemize}
\item
  Interesting individual case studies:

@@

\subsection{Comparison with archaeobotanical
data}\label{comparison-with-archaeobotanical-data}

The failure of our hindcast models to predict the occurrence of species
in archaeobotanical assemblages has several possible explanations. Since
they do accurately predict the test dataset , a likely culprit is
overfitting of the models to the present environment. This implies that
the modelled palaeodistributions should be seen as conservative
estimates or a minimal range. Another obvious flaw in our methodology is
that the time slices used for palaeoclimatic reconstruction are very
broad---each covering around two millennia---and therefore potentially
unrepresentative of the environment around sites at the specific time at
which they were occupied. The variable quality of the archaeological
test dataset, especially in terms of chronology, is also a plausible
factor.

At the same time, we cannot rule out more substantive reasons for the
discrepency between predicted and observed archaeological occurrences.
The niches of the modelled species could have changed since the Early
Holocene, which would not be captured in a model trained purely on
modern specimens. Human economic choices---mobility, foraging
strategies, cultivation, etc.---could also produce archaeobotanical
assemblages whose composition depart significantly from that of the
surrounding local flora. Further refinement of the methodology for
hindcast palaeoecological niche models, for example using more finely
resolved palaeoclimate sequences \citep{KargerEtAl2023}, hyperparameter
tuning to avoid overfitting, and improved archaeological datasets, would
help disentangle these potential explanations.

@@

\section{References}\label{references}

\begin{center}\rule{0.5\linewidth}{0.5pt}\end{center}

  \bibliography{references.bib}
